One page description

### 2016-09-30 publications/Journal/2-background.tex
Consistent italics and sans-serif fonts
First 3 sections review for style and acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

This section briefly introduces the synchronization protocols used by dxex: conservative and optimistic synchronization.

@@

Rolling back simulation time requires the node to store past model states, such that they can be restored later.
All incoming and outgoing events need to be stored as well.
Incoming events are injected again after a rollback, when their time has been reached again.
Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
In practice, a single causality error can have significant repercussions on performance.

@@

The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
Apart from overhead in CPU time, a significant memory overhead is present: intermediate states are stored up to a point where it can be considered \textit{irreversible}.
Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.
Instead of depending on the theoretically defined safe lookahead, performance is related to the actually perceived lookahead.

>>>>

This section briefly introduces the synchronization protocols used by \textit{dxex}: conservative and optimistic synchronization.

@@

Rolling back simulation time requires the node to store past model states, such that they can be restored later.
All incoming and outgoing events need to be stored as well.
Incoming events are injected again after a rollback, when their time has been reached again.
Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
Cancelling events can cause further rollbacks as the receiving node might also have to roll back its state.
In practice a single causality error can have significant repercussions on performance.

@@

The main advantage is that a small lookahead, caused by a very infrequent event, will not limit performance.
%The main advantage is that performance is not limited by a small lookahead caused by a very infrequent event.
If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
Apart from overhead in CPU time, a significant memory overhead is present: intermediate states are stored up to a point where it can be considered \textit{irreversible}.
Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.
Instead of depending on the theoretically defined safe lookahead, performance is related to the actually perceived lookahead.

### 2016-09-30 publications/Journal/3-features.tex
First 3 sections review for style and acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local (using \textit{tcmalloc}) memory pools.
Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.

@@

Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise. Providing this default implementation has no impact in sequential or optimistic simulation, since the function is never called the compiler will optimize it out. By providing this default implementation in the model base class we ensure that a model can run in sequential, optimistic and conservative simulation by default.

@@

Furthermore, allocating and deallocating memory by making calls to the operating system, as typically done by calls such as \texttt{malloc}, happens sequentially.
To counter this, our memory allocators are backed by a thread-aware pooling library.
In a sequential simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.

%TODO what does it mean to "pool aggressively?"
Intra-kernel events are pooled aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
A simulation with many inter kernel events suffers a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators.

@@

\subsubsection{Random Number Generators}
Random Number Generators (RNG) are another aspect of the program that results in sequentialization.
All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{Random}.
As such, a non-trivial amount of time in a simulation is often spent waiting for an RNG.

@@

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default STL random number generator.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

>>>>

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local %(using \textit{tcmalloc})% see page 6 for introducing tcmalloc
 memory pools.
Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.

@@

Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default value $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise. Providing this default implementation has no impact in sequential or optimistic simulation, since the function is never called the compiler will optimize it out. By providing this default implementation in the model base class we ensure that a model can run in sequential, optimistic and conservative simulation by default.

@@

Furthermore, allocating and deallocating memory by making calls to the operating system, as is typically done by calls such as \texttt{malloc}, happens sequentially.
To counter this, our memory allocators are backed by a thread-aware pooling library.
In a sequential simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.

Intra-kernel events can be (de)allocated without requiring synchronization with the other kernels. 
They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation.
In contrast inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
As a result simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena ~\cite{Arena} allocators.

@@

\subsubsection{Random Number Generators}
Random Number Generators (RNG) are another aspect of the program that can limit parallelization.
All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{Random}.
As such, a non-trivial amount of time in a simulation is often spent waiting for an RNG.

@@

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.
